---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rix: Reproducible Environments with Nix

## Installation

You can install the development version of rix from
[GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("b-rodrigues/rix")
```

If you are on Windows, you need the windows subsystem for linux 2 (WSL2) to
run Nixpkgs.

## Introduction

<!-- badges: start -->
[![R-CMD-check](https://github.com/b-rodrigues/rix/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/b-rodrigues/rix/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{rix}` is an R package that provides functions to help you setup reproducible
development environments that contain all the required packages that you need
for your project. This is achieved by using the Nix package manager that you
must install separately. The Nix package manager is extremely powerful: with it,
it is possible to work on totally reproducible development environments, and
even install old releases of R and R packages. With Nix, it is essentially
possible to replace `{renv}` and Docker combined. If you need other tools or
languages like Python or Julia, this can also be done easily. Nix is available
for Linux, macOS and Windows.

## The Nix package manager

Nix is a piece of software that can be installed on your computer (regardless of
OS) and can be used to install software like with any other package manager. If
you're familiar with the Ubuntu Linux distribution, you likely have used
`apt-get` to install software. On macOS, you may have used `homebrew` for
similar purposes. Nix functions in a similar way, but has many advantages over
classic package managers. The main advantage of Nix, at least for our purposes,
is that its repository of software is huge. As of writing, it contains more than
80'000 packages, and the entirety of CRAN is available through Nix's
repositories. This means that using Nix, it is possible to install not only R,
but also all the packages required for your project. The obvious question is why
use Nix instead of simply installing R and R packages as usual. The answer is
that Nix makes sure to install every dependency of any package, up to required
system libraries. For example, the `{xlsx}` package requires the Java
programming language to be installed on your computer to successfully install.
This can be difficult to achieve, and `{xlsx}` bullied many R developers
throughout the years (especially those using a Linux distribution, `sudo R CMD
javareconf` still plagues my nightmares). But with Nix, it suffices to declare
that we want the `{xlsx}` package for our project, and Nix figures out
automatically that Java is required and installs and configures it. It all just
happens without any required intervention from the user. The second advantage of
Nix is that it is possible to *pin* a certain *revision* for our project.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
installed.

## Rix workflow

The idea of `{rix}` is for you to declare the environment you need, using  the
provided `rix()` function, which in turn generates the required file for Nix
to actually generate that environment. You can then use this environment to 
either work interactively, or run R scripts. It is possible to have as many
environments as projects. Each environment is isolated (or not, it's up to you).

The main function of `{rix}` is called `rix()`. `rix()` has several arguments:

- the R version you need for your project
- a list of R packages that your project needs
- an optional list of additional software (for example, a Python interpreter, or Quarto)
- an optional list with packages to install from Github
- whether you want to use RStudio as an IDE for your project (or VS Code, or another environment)
- a path to save a file called `default.nix`.

For example:

```{r, eval = FALSE}
rix(r_ver = "current", pkgs = c("dplyr", "chronicler"), ide = "rstudio")
```

The call above writes a `default.nix` file in the current working directory. This
`default.nix` can in turn be used by Nix to build an environment containing
RStudio, the current (or latest) version of R, and the latest versions of the
`{dplyr}` and `{chronicler}` packages. In th case of RStudio, it actually needs
to be installed for each environment. This is because RStudio changes some
default environment variables and a globally installed RStudio (the one you
install normally) would not recognize the R installed in the Nix environment.
This is not the case for other IDEs such as VS code or Emacs. Another example:

```{r, eval = FALSE}
rix(r_ver = "4.1.0", pkgs = c("dplyr", "chronicler"), ide = "code")
```

This call will generate a `default.nix` that installs R version 4.1.0, with the `{dplyr}`
and `{chronicler}` packages. Because the user wishes to use VS Code, the `ide` argument
was set to "code". This installs the required `{languageserver}` package as well, but unlike
`ide = "rstudio"` does not install VS Code in that environment. Users should instead use 
the globally installed VS Code.

### default.nix

The Nix package manager can be used to build reproducible development
environments according to the specifications found in a file called
`default.nix`. To make it easier for R programmers to use Nix, `{rix}` can be
used to write this file for you. Once this file has been written, go to where
you chose to write it (ideally in a new, empty folder that will be the root
folder of your project) and use the Nix package manager to build the environment.
Call the following function in a terminal:

```
nix-build
```

Once Nix done building the environment, you can start working on it interactively
by using the following command:

```
nix-shell
```

You will *drop* in a Nix shell. You can now call the IDE of your choice. For
RStudio, simply call:

```
rstudio
```

This will start RStudio. RStudio will use the version of R and library of packages
from that environment.

### Running programs from an environment

You could create a bash script that you put in the path to make this process
more streamlined. For example, if your project is called `housing`, you could
create this script and execute it to start your project:

```
!#/bin/bash
nix-shell /absolute/path/to/housing/default.nix --run rstudio
```

This will execute RStudio in the environment for the `housing` project. If you
use `{targets}` you could execute the pipeline in the environment by running:

```
nix-shell /absolute/path/to/housing/default.nix --run Rscript -e 'targets::tar_make()'
```

It's possible to execute the pipeline automatically using a so-called "shell
hook". See the "Non-interactive use" vignette for more details.

## Contributing

This package is developed using the `{fusen}` package. If you want to
contribute, please edit the `.Rmd` files found in the `dev/` folder. Then,
inflate the package using `fusen::inflate_all()`. If no errors are found
(warning and notes are ok), then commit and open a PR. To learn how to use
`{fusen}` (don't worry, it's super easy), refer to this
[vignette](https://thinkr-open.github.io/fusen/articles/How-to-use-fusen.html).

