---
title: "Development actions history"
output: html_document
editor_options: 
  chunk_output_type: console
---

All commands that you use to use when developing packages...

# First time just after creating the project

- Fill the following chunk to create the DESCRIPTION of your package

```{r description, eval=FALSE}
# Describe your package
fusen::fill_description(
  pkg = here::here(),
  fields = list(
    Title = "Rix: Reproducible Environments with Nix",
    URL = "https://b-rodrigues.github.io/rix/",
    BugReports = "https://github.com/b-rodrigues/rix",
    Description = "Provides helper functions to create reproducible development environments using the Nix package manager.",
    Version = "0.0.1",
    `Authors@R` = c(
      person("Bruno", "Rodrigues", email = "bruno@brodrigues.co", role = c("aut", "cre"), comment = c(ORCID = "0000-0002-3211-3689")),
      person("Philipp", "Baumann", email = "baumann-philipp@protonmail.com", role = "aut", comment = c(ORCID = "0000-0002-3194-8975"))
    )
  ),
  overwrite = TRUE
)
# Define License with use_*_license()
usethis::use_gpl3_license()

```

Add codemeta file:

```{r, eval = F}
codemetar::write_codemeta()

```

# All-in-one function to deploy publicly on GitHub

Either use this GitHub all-in-one function or choose the steps one-by-one in the following sections.
See `vignette("share-on-a-github-website", package = "fusen")` for more information.

```{r, eval=FALSE}
# _All-in-one share of your package and its documentation on GitHub
#fusen::init_share_on_github()
```


# Start using git

```{r, eval=FALSE}
usethis::use_git()
# Deal with classical files to ignore
#usethis::git_vaccinate()
# Use main for primary branch
#usethis::git_default_branch_rename()
```

# Set extra sources of documentation

```{r, eval=FALSE}
# Install a first time
#remotes::install_local()
# README
#usethis::use_readme_rmd()
# Code of Conduct
#usethis::use_code_of_conduct("contact@fake.com")
# NEWS
#usethis::use_news_md()
```

**From now, you will need to "inflate" your package at least once to be able to use the following commands. Let's go to your flat template, and come back here later if/when needed.**

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_save_r_nix_revs.Rmd",
               vignette_name = "z - Developers Vignette: Save the Nix Package Versions data",
               overwrite = TRUE)
```

```{r}
local({
  knitr::knit("vignettes/z-developers-vignette-save-the-nix-package-versions-data.Rmd",
              output = tempfile())
  usethis::use_data(r_nix_revs,
                    overwrite = TRUE)
})

```

```{r}
fusen::inflate(flat_file = "dev/flat_data_doc.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_build_envs.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_cicd.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```


```{r}
fusen::inflate(flat_file = "dev/flat_zzz.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_get_os.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_cran_archive.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

# Vignettes for users

```{r}
fusen::inflate(flat_file = "dev/1-getting_started.Rmd",
               vignette_name = "a - Getting started",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/2a-linux_win.Rmd",
               vignette_name = "b1 - Setting up and using rix on Linux and Windows",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/2b-macos.Rmd",
               vignette_name = "b2 - Setting up and using rix on macOS",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/3-building_envs_with_rix.Rmd",
               vignette_name = "c - Using rix to build project specific environments",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/4a-install_r_pkgs.Rmd",
               vignette_name = "d1 - Installing R packages in a Nix environment",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/4b-install_sys_pkgs.Rmd",
               vignette_name = "d2 - Installing system tools and TexLive packages in a Nix environment",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/5-interactive_use.Rmd",
               vignette_name = "e - Interactive use",
               overwrite = TRUE)
```


```{r}
fusen::inflate(flat_file = "dev/literate_programming.Rmd",
               vignette_name = "z - Advanced topic: Building an environment for literate programming",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/pkgs_with_remotes.Rmd",
               vignette_name = "z - Advanced topic: Handling packages with remote dependencies",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/raps_with_nix.Rmd",
               vignette_name = "z - Advanced topic: Reproducible Analytical Pipelines with Nix",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/subshells.Rmd",
               vignette_name = "z - Advanced topic: Running R or Shell Code in Nix from R",
               overwrite = TRUE)
```

# Generating the inst/extdata/default.nix

```{r, eval = FALSE}
rix(r_ver = "976fa3369d722e76f37c77493d99829540d43845",
    r_pkgs = NULL,
    system_pkgs = NULL,
    git_pkgs = list(
      package_name = "rix",
      repo_url = "https://github.com/b-rodrigues/rix/",
      branch_name = "master",
      commit = "ae39d2142461688b1be41db800752a949ebb3c7b"
    ),
    ide = "other",
    project_path = "inst/extdata",
    overwrite = TRUE)
```

Here's another, for testing purposes:

```{r, eval = FALSE}
rix(r_ver = "8ad5e8132c5dcf977e308e7bf5517cc6cc0bf7d8",
    r_pkgs = c("dplyr", "AER@1.2-8"),
    system_pkgs = c("quarto"),
    git_pkgs = list(
      package_name = "rix",
      repo_url = "https://github.com/b-rodrigues/rix/",
      branch_name = "master",
      commit = "da581f90273cb1ccfedbe18808097bf33b84d63c"
    ),
    ide = "other",
    project_path = "inst/extdata",
    overwrite = TRUE,
    print = TRUE)
```


# Package development tools
## Use once

```{r, eval=FALSE}
# Pipe
usethis::use_pipe()

# package-level documentation
usethis::use_package_doc()

# GitHub
# Add your credentials for GitHub
usethis::create_github_token()
gitcreds::gitcreds_set()

# _Or one step at a time
# Send your project to a new GitHub project (included in `init_share_on_github()`)
usethis::use_github()

# Set Continuous Integration
# _GitHub (included in `init_share_on_github()`)
usethis::use_github_action_check_standard()
usethis::use_github_action("pkgdown")
usethis::use_github_action("test-coverage")
# _GitLab
gitlabr::use_gitlab_ci(type = "check-coverage-pkgdown")
```

## Use everytime needed

```{r}
# Simulate package installation
pkgload::load_all()

# Generate documentation and deal with dependencies
attachment::att_amend_desc()

# Check the package
devtools::check()

# Add a new flat template
fusen::add_flat_template("add")
```

# Share the package

```{r}
# set and try pkgdown documentation website locally
usethis::use_pkgdown()
pkgdown::build_site()

# build the tar.gz with vignettes to share with others
devtools::build(vignettes = TRUE)

# Share your package and its documentation on GitHub
# usethis::create_github_token()
# gitcreds::gitcreds_set()
fusen::init_share_on_github()
```

