---
title: "Get SRI hash and dependencies"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r development, include=FALSE}
library(testthat)
```


`get_sri_hash_deps()` returns the SRI hash of a NAR serialized path to a cloned
Github repository, or package source downloaded from the CRAN archives, alongside
that packages' build dependencies. These hashes are used by Nix for security purposes. In
order to get the hash, a GET to a service I've made gets made. This request
gets handled by a server with Nix installed, and so the SRI hash can get computed`
by `nix hash path --sri path_to_repo`.

```{r function-get_sri_hash_deps}
#' get_sri_hash_deps Get the SRI hash of the NAR serialization of a Github repo
#' @param repo_url A character. The URL to the package's Github repository or to the `.tar.gz` package hosted on CRAN.
#' @param branch_name A character. The branch of interest, NULL for archived CRAN packages.
#' @param commit A character. The commit hash of interest, for reproducibility's sake, NULL for archived CRAN packages.
#' @importFrom httr content GET http_error
#' @return The SRI hash as a character
get_sri_hash_deps <- function(repo_url, branch_name, commit){
  result <- httr::GET(paste0("http://git2nixsha.dev:1506/hash?repo_url=",
                             repo_url,
                             "&branchName=",
                             branch_name,
                             "&commit=",
                             commit))

  if(http_error(result)){
    stop(paste0("Error in pulling URL: ", repo_url, ". If it's a Github repo, check the url, branch name and commit. Are these correct? If it's an archived CRAN package, check the name of the package and the version number."))
  }


  lapply(httr::content(result), unlist)

}
```

```{r tests-get_sri_hash_deps}
testthat::test_that("get_sri_hash_deps returns correct sri hash and dependencies of R packages", {
  testthat::expect_equal(
              get_sri_hash_deps("https://github.com/rap4all/housing/",
                                "fusen",
                                "1c860959310b80e67c41f7bbdc3e84cef00df18e"),
              list(
                "sri_hash" = "sha256-s4KGtfKQ7hL0sfDhGb4BpBpspfefBN6hf+XlslqyEn4=",
                "deps" = "dplyr ggplot2 janitor purrr readxl rlang rvest stringr tidyr"
              )
            )
})

testthat::test_that("Internet is out for fetchgit()", {

  testthat::local_mocked_bindings(
    http_error = function(...) TRUE
    )

  expect_error(
    get_sri_hash_deps(
      "https://github.com/rap4all/housing/",
      "fusen",
      "1c860959310b80e67c41f7bbdc3e84cef00df18e"
    ),
    'Error in pulling',
    )

})

```
