---
title: "2a - Setting up and using rix on Linux and Windows"
output: html_document
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 80
---

*This vignette will discuss Linux and Windows-specific topics. If you're not
using either of these systems, you can ignore this vignette, and read the
[Setting up and using rix on macOS]() vignette instead.*

## Introduction

When it comes to Nix, there are really only two supported operating systems:
macOS and Linux distributions. Windows is "supported" because it is actually
running Linux thanks to WSL2. In practice this means that Linux distributions
and Windows can be considered one system, and macOS another, separate, system.
Because Windows is really running Linux under the hood thanks to WSL2, this
means that WSL2 needs to be running on your Windows system before you attempt to
install Nix. But it is important to know that you can run `{rix}` even if you
don't have Nix installed, which means you can generate Nix expressions, you
just can't build them.

### Windows pre-requisites

If you are on Windows, you need the Windows Subsystem for Linux 2 (WSL2) to run
Nix. If you are on a recent version of Windows 10 or 11, you can simply run this
as an administrator in PowerShell:

```
wsl --install
```

You can find further installation notes at [this official MS
documentation](https://learn.microsoft.com/en-us/windows/wsl/install).

We recommend to activate systemd in Ubuntu WSL2, mainly because this supports 
other users than `root` running Nix. To set this up, please do as outlined
[this official Ubuntu blog entry](https://ubuntu.com/blog/ubuntu-wsl-enable-systemd):

```sh
# in WSL2 Ubuntu shell
sudo -i
nano /etc/wsl.conf
# add this entry
[boot]
systemd=true
# then restart running instance from PowerShell
wsl --shutdown
# relaunch Ubuntu WSL2
```

Afterwards, you can install Nix like business as usual. You can proceed with the
Determinate Systems installer.

## Installing Nix

You can use `{rix}` to generate Nix expressions even if you don't have Nix
installed on your system, but obviously, you need to install Nix if you actually
want to build the defined development environment and use them. Installing (and
uninstalling) Nix is quite simple, thanks to the installer from [Determinate
Systems](https://determinate.systems/posts/determinate-nix-installer), a company
that provides services and tools built on Nix. Simply open a terminal and run
the following line (on Windows, if you cannot or have decided not to activate
systemd, then you have to append `--init none` to the command. More details
about this you can find at [The Determinante Nix
Installer](https://github.com/DeterminateSystems/nix-installer)):

```{sh, eval=FALSE}
curl --proto '=https' --tlsv1.2 -sSf \
    -L https://install.determinate.systems/nix | \
     sh -s -- install
```

Once you have Nix installed, you can build the expressions you generate with `{rix}`!

## What if you don't have R already installed?

If you have successfully installed Nix, but don't have yet R installed on your
system, you could install R as you would usually do on your operating system,
and then install the `{rix}` package, and from there, generated project-specific
expressions and build them. But you could also install R using Nix. Running the
following line in a terminal will drop you in an interactive R session that you
can use to start generating expressions:

```
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/b-rodrigues/rix/master/inst/extdata/default.nix)"
```

This should immediately start an R session inside your terminal. You can now run
something like this:

```
rix(r_ver = "latest",
    r_pkgs = c("dplyr", "ggplot2"),
    system_pkgs = NULL,
    git_pkgs = NULL,
    ide = "other",
    project_path = ".",
    overwrite = TRUE)
```

to generate a `default.nix`, and then use that file to generate an environment
with R, `{dplyr}` and `{ggplot2}`. If you need to add packages for your
project, rerun the command above, but add the needed packages to `r_pkgs`.

## Generating expressions

Once you have R installed, either through the usual installer for your operating
system, or through Nix as explained previously, you can now start building
project-specific development environments.

Start an R session, and install `{rix}` if that's not already done. Because
`{rix}` is not yet on CRAN, the easiest way is to install it from its
r-universe:

```{r, eval=FALSE}
install.packages("rix", repos = c("https://b-rodrigues.r-universe.dev",
  "https://cloud.r-project.org"))
```

You can then use the `{rix}` package to generate expressions. For example, here
is how to generate an expression that installs version 4.2.1 of R and the
`{dplyr}` and `{chronicler}` packages:

```{r, eval=FALSE}
library(rix)

rix(r_ver = "4.2.1",
    r_pkgs = c("dplyr", "chronicler"),
    ide = "other")
```

If you need to install packages from GitHub or other tools, refer to the
vignette [installing R packages in a Nix environment]() to learn how to install
packages from Git or older packages from the CRAN archives, and also refer to
the vignette [installing system tools and TexLive packages in a Nix
environment]() to learn how to add tools other programs to your environment such
as [Quarto](https://quarto.org/) and TexLive packages for writing documents
using the LaTeX typesetting system.

