---
title: "b1 - Setting up and using rix on Linux and Windows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{b1-setting-up-and-using-rix-on-linux-and-windows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(rix)
```

*This vignette will discuss Linux and Windows-specific topics. If you're not
using either of these systems, you can ignore this vignette, and read the
`vignette("b2-setting-up-and-using-rix-on-macos")`
vignette instead.*


## Introduction

When it comes to Nix, there are really only two supported operating systems:
macOS and Linux distributions. Windows is "supported" because it is actually
running Linux thanks to WSL2. In practice this means that Linux distributions
and Windows can be considered one system, and macOS another, separate, system.
Because Windows is really running Linux under the hood thanks to WSL2, this
means that WSL2 needs to be running on your Windows system before you attempt to
install Nix. But it is important to know that you can run `{rix}` even if you
don't have Nix installed, which means you can generate Nix expressions, you
just can't build them. So if you can’t install Nix on your system, but have 
R already installed, you can skip to the last section of this vignette to 
simply install the `{rix}` package.


## Installing Nix

### Windows pre-requisites

If you are on Windows, you need the Windows Subsystem for Linux 2 (WSL2) to run
Nix. If you are on a recent version of Windows 10 or 11, you can simply run this
as an administrator in PowerShell:

```
wsl --install
```

You can find further installation notes at [this official MS
documentation](https://learn.microsoft.com/en-us/windows/wsl/install).

We recommend to activate systemd in Ubuntu WSL2, mainly because this supports 
other users than `root` running Nix. To set this up, please do as outlined
[this official Ubuntu blog entry](https://ubuntu.com/blog/ubuntu-wsl-enable-systemd):

```sh

# in WSL2 Ubuntu shell

sudo -i
nano /etc/wsl.conf
```

This will open the `/etc/wsl.conf` in a nano, a command
line text editor. Add the following line:

```
[boot]
systemd=true
```

Save the file with CTRL-O and then quit nano with CTRL-X.
Then, type the following line in powershell:

```
wsl --shutdown
```

and then relaunch WSL (Ubuntu) from the start menu.

Afterwards, you can install Nix like business as usual. You can proceed with the
Determinate Systems installer.

### Using the Determinate Systems installer

You can use `{rix}` to generate Nix expressions even if you don't have Nix
installed on your system, but obviously, you need to install Nix if you actually
want to build the defined development environment and use them. 

Installing (and
uninstalling) Nix is quite simple, thanks to the installer from [Determinate
Systems](https://determinate.systems/posts/determinate-nix-installer), a company
that provides services and tools built on Nix.

Do not use your operating system’s package manager to install Nix. Instead, 
simply open a terminal and run the following line (on Windows, if you cannot or 
have decided not to activate systemd, then you have to append `--init none`
to the command. You can find more details about this on [The Determinate Nix
Installer page](https://github.com/DeterminateSystems/nix-installer)):


```{sh parsermd-chunk-1, eval = FALSE}
curl --proto '=https' --tlsv1.2 -sSf \
    -L https://install.determinate.systems/nix | \
     sh -s -- install
```

Once you have Nix installed, you can build the expressions you generate with `{rix}`!

On Linux, once Nix is installed, all the software that will be installed through
Nix will be saved to the `/nix` directory on the root partition. It is common
for Linux users to have a separate partition for `/`, which may be small.
Complete development environments built with Nix can take up much space, so if
the available space on your root partition is limited, we advise you to mount the
`/nix` folder on another partition with more space (for example, a secondary
hard drive). For this, edit `/etc/fstab` and add the following line at the end:

```
/home/path_to/nix /nix none bind 0 0
```

This will map `/nix` to `/home/path_to/nix` which can be on a larger partition.
If you have enough space on your root partition, you can ignore the above
instructions.


## Case 1: you don’t have R installed and wish to install it using Nix as well

If you have successfully installed Nix, but don't have yet R installed on your
system, you could install R as you would usually do on your operating system,
and then install the `{rix}` package, and from there, generated project-specific
expressions and build them. But you could also install R using Nix. Running the
following line in a terminal will drop you in an interactive R session that you
can use to start generating expressions:

```
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/master/inst/extdata/default.nix)"
```

This should immediately start an R session inside your terminal. You can now run
something like this:

```{r, eval = F}
rix(
  r_ver = "latest",
  r_pkgs = c("dplyr", "ggplot2"),
  system_pkgs = NULL,
  git_pkgs = NULL,
  ide = "other",
  project_path = ".",
  overwrite = TRUE
)
```

to generate a `default.nix`, and then use that file to generate an environment
with R, `{dplyr}` and `{ggplot2}`. If you need to add packages for your project,
rerun the command above, but add the needed packages to `r_pkgs`. Beware that if
your already have a `default.nix` file in the working directory, running
`rix()` with the `overwrite = TRUE` argument will overwrite it! So make sure
that you are using a version control system for your projects to avoid bad surprises.

More details about managing project-specific `default.nix` are detailled in the 
vignette `vignette("d1-installing-r-packages-in-a-nix-environment")` and
`vignette("d2-installing-system-tools-and-texlive-packages-in-a-nix-environment")`.

You could also include `{rix}` in your project-specific environments, by
generating a `default.nix` like so:

```{r, eval = F}
rix(
  r_ver = "latest",
  r_pkgs = NULL,
  git_pkgs = list(
    package_name = "rix",
    repo_url = "https://github.com/ropensci/rix",
    commit = "76d1bdd03d78589d399b4b9d473ecde616920a82"
  ),
  ide = "other",
  project_path = ".",
  overwrite = TRUE
)
```

Change the commit to a more recent one and adapt the `project_path` argument if needed.


## Case 2: you have R installed through your OS’s package manager

If you have installed R on your system through the usual means of installation
(so not Nix, as described before), you can install the `{rix}` package as usual
as well. To install `{rix}`, run:

```{r parsermd-chunk-2-0, eval = FALSE}
install.packages("rix")
```

or use r-universe:

```{r parsermd-chunk-2, eval = FALSE}
install.packages("rix", repos = c(
  "https://ropensci.r-universe.dev",
  "https://cloud.r-project.org"
))
```

You can then use the `{rix}` package to generate expressions. Consult the 
next vignette `vignette("c-using-rix-to-build-project-specific-environments")` to learn more.

