---
title: "d1 - Installing R and Python packages in a Nix environment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{d1-installing-r-packages-in-a-nix-environment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(rix)
```

## Introduction

You now know how to declare and build reproducible development environments
using `{rix}` and Nix. This vignette will explain how to install specific
versions of CRAN packages and how to install packages from GitHub.

## A word of caution

It is important at this stage to understand that you should not call
`install.packages()` from a running Nix environment. Doing so will raise an
error to avoid issues. If you want to add packages while analyzing data, you
need to add it the `default.nix` expression and rebuild the environment. The
same goes for installing packages from GitHub; use the method described in this
vignette instead of using something like `remotes::install_github()`.

We recommend you create a script called `create_env.R` or similar, and add the call to `rix()` there:

```
library(rix)

rix(r_ver = "4.4.0",
    r_pkgs = c("dplyr", "ggplot2"),
    system_pkgs = NULL,
    git_pkgs = NULL,
    ide = "code",
    project_path = path_default_nix,
    overwrite = TRUE,
    print = TRUE)
```

Then, add the packages you need to `r_pkgs` and run the script again. Then, build
the environment using `nix-build` again, and drop into it using `nix-shell`.
Calling `install.packages()` is a bad idea for several reasons:

- it goes against the idea of defining an environment in a declarative way. If
  you were able to add packages using `install.packages()`, your environment
  would end up in a state where the `default.nix` definition of the environment
  and the actual environment don't match anymore.
- using `install.packages()` would likely simply not work, and if it would work,
  it would cause issues. For example, if you call `install.packages("ggplot2")`
  from one Nix shell, it will not install `{ggplot2}` "inside" the Nix shell,
  but will install it on your user's system library of packages. This is because
  the Nix shell cannot be changed at run-time, and so, R will instead install
  the packages in the user's library. This version of `{ggplot2}`, because it is
  in that system-wide library of packages, will be available to any other Nix
  shells. If you call `install.packages("ggplot2")` again from another Nix shell,
  say 6 months later, this will replace the first version of `{ggplot2}` with
  the latest version.

Ideally, you should only manage R versions and R packages using Nix, and
uninstall any system-managed version of R and R packages. But if you do wish to
keep a system-managed version of R and R packages, `rix::rix()` also runs
`rix::rix_init()` automatically which generates an `.Rprofile` file that avoids
any clashes between your global user library and Nix-managed libraries of R
packages.

## Installing old packages archived on CRAN

It is possible to install an arbitrary version of a package that has been
archived on CRAN:

```{r, eval = F}
path_default_nix <- tempdir()

rix(
  r_ver = "4.2.1",
  r_pkgs = c("dplyr@0.8.0", "janitor@1.0.0"),
  system_pkgs = NULL,
  git_pkgs = NULL,
  ide = "none",
  project_path = path_default_nix,
  overwrite = TRUE
)
```

```
#> let
#>  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a.tar.gz") {};
#>
#>   git_archive_pkgs = [
#>     (pkgs.rPackages.buildRPackage {
#>       name = "dplyr";
#>       src = pkgs.fetchzip {
#>        url = "https://cran.r-project.org/src/contrib/Archive/dplyr/dplyr_0.8.0.tar.gz";
#>        sha256 = "sha256-f30raalLd9KoZKZSxeTN71PG6BczXRIiP6g7EZeH09U=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages)
#>           assertthat
#>           glue
#>           magrittr
#>           pkgconfig
#>           R6
#>           Rcpp
#>           rlang
#>           tibble
#>           tidyselect
#>           BH
#>           plogr;
#>       };
#>     })
#>
#>     (pkgs.rPackages.buildRPackage {
#>       name = "janitor";
#>       src = pkgs.fetchzip {
#>        url = "https://cran.r-project.org/src/contrib/Archive/janitor/janitor_1.0.0.tar.gz";
#>        sha256 = "sha256-3NJomE/CNbOZ+ohuVZJWe2n1RVGUm8x8a0A0qzLmdN4=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages)
#>           dplyr
#>           tidyr
#>           snakecase
#>           magrittr
#>           purrr
#>           rlang;
#>       };
#>     })
#>   ];
#>
#>   system_packages = builtins.attrValues {
#>     inherit (pkgs)
#>       R
#>       glibcLocales
#>       nix;
#>   };
#>
#> in
#>
#> pkgs.mkShell {
#>   LOCALE_ARCHIVE = if pkgs.system == "x86_64-linux" then  "${pkgs.glibcLocales}/lib/locale/locale-archive" else "";
#>   LANG = "en_US.UTF-8";
#>    LC_ALL = "en_US.UTF-8";
#>    LC_TIME = "en_US.UTF-8";
#>    LC_MONETARY = "en_US.UTF-8";
#>    LC_PAPER = "en_US.UTF-8";
#>    LC_MEASUREMENT = "en_US.UTF-8";
#>
#>   buildInputs = [ git_archive_pkgs   system_packages   ];
#>
#> }
```


The above expression will install R version 4.2.1, and `{dplyr}` at version
0.8.0 and `{janitor}` at version 1.0.0. This can be useful, especially for
packages that have been archived, but otherwise, this feature should ideally be
used sparingly, for two reasons. First, if you want to reconstruct an
environment as it was around 2019, use the version of R that was current at that
time using the `date` argument in `rix()`. This will ensure that every package
that gets installed is at a version compatible with that version of R, which
might not be the case if you need to install a very old version of one
particular package. Second, doing so will install the package from source. For
packages that don't require compilation, this should be fine, but packages that
require compilation will likely fail to compile successfully. We are working on
handling this better for future versions of `{rix}`.

## Installing packages from GitHub

It is also possible to install packages from GitHub:

```{r, eval = F}
path_default_nix <- tempdir()

rix(
  r_ver = "4.2.1",
  r_pkgs = c("dplyr", "janitor"),
  git_pkgs = list(
    list(
      package_name = "housing",
      repo_url = "https://github.com/rap4all/housing/",
      commit = "1c860959310b80e67c41f7bbdc3e84cef00df18e"
    ),
    list(
      package_name = "fusen",
      repo_url = "https://github.com/ThinkR-open/fusen",
      commit = "d617172447d2947efb20ad6a4463742b8a5d79dc"
    )
  ),
  ide = "none",
  project_path = path_default_nix,
  overwrite = TRUE
)
```

```{r parsermd-chunk-4, echo = F}
#> # This file was generated by the {rix} R package v0.7.1 on 2024-07-01
#> # with following call:
#> # >rix(r_ver = "79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a",
#> #  > r_pkgs = c("dplyr",
#> #  > "janitor"),
#> #  > git_pkgs = list(list(package_name = "housing",
#> #  > repo_url = "https://github.com/rap4all/housing/",
#> #  > commit = "1c860959310b80e67c41f7bbdc3e84cef00df18e"),
#> #  > list(package_name = "fusen",
#> #  > repo_url = "https://github.com/ThinkR-open/fusen",
#> #  > commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")),
#> #  > ide = "none",
#> #  > project_path = path_default_nix,
#> #  > overwrite = TRUE)
#> # It uses nixpkgs' revision 79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a for reproducibility purposes
#> # which will install R version 4.2.1.
#> # Report any issues to https://github.com/ropensci/rix
#> let
#>  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a.tar.gz") {};
#>
#>   rpkgs = builtins.attrValues {
#>     inherit (pkgs.rPackages)
#>       dplyr
#>       janitor;
#>   };
#>
#>   git_archive_pkgs = [
#>     (pkgs.rPackages.buildRPackage {
#>       name = "housing";
#>       src = pkgs.fetchgit {
#>         url = "https://github.com/rap4all/housing/";
#>         rev = "1c860959310b80e67c41f7bbdc3e84cef00df18e";
#>         sha256 = "sha256-s4KGtfKQ7hL0sfDhGb4BpBpspfefBN6hf+XlslqyEn4=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages)
#>           dplyr
#>           ggplot2
#>           janitor
#>           purrr
#>           readxl
#>           rlang
#>           rvest
#>           stringr
#>           tidyr;
#>       };
#>     })
#>
#>
#>     (pkgs.rPackages.buildRPackage {
#>       name = "fusen";
#>       src = pkgs.fetchgit {
#>         url = "https://github.com/ThinkR-open/fusen";
#>         rev = "d617172447d2947efb20ad6a4463742b8a5d79dc";
#>         sha256 = "sha256-TOHA1ymLUSgZMYIA1a2yvuv0799svaDOl3zOhNRxcmw=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages)
#>           attachment
#>           cli
#>           desc
#>           devtools
#>           glue
#>           here
#>           magrittr
#>           parsermd
#>           roxygen2
#>           stringi
#>           tibble
#>           tidyr
#>           usethis
#>           yaml;
#>       };
#>     })
#>    ];
#>
#>   system_packages = builtins.attrValues {
#>     inherit (pkgs)
#>       R
#>       glibcLocales
#>       nix;
#>   };
#>
#> in
#>
#> pkgs.mkShell {
#>   LOCALE_ARCHIVE = if pkgs.system == "x86_64-linux" then  "${pkgs.glibcLocales}/lib/locale/locale-archive" else "";
#>   LANG = "en_US.UTF-8";
#>    LC_ALL = "en_US.UTF-8";
#>    LC_TIME = "en_US.UTF-8";
#>    LC_MONETARY = "en_US.UTF-8";
#>    LC_PAPER = "en_US.UTF-8";
#>    LC_MEASUREMENT = "en_US.UTF-8";
#>
#>   buildInputs = [ git_archive_pkgs rpkgs  system_packages   ];
#>
#> }
```

This will install two packages from GitHub: the `{housing}` package and more
specifically the code as it is in the `fusen` branch; however the branch name is
not required, as the commit is enough to pin the exact version of the code
needed. The `{fusen}` package is also installed, as of commit `d617172447d`.

If you want to install a package from GitHub, which store the R package in a subfolder,
you should specify the subfolder in the `repo_url` argument. For example, if you want to
install the R version of the package `{BPCells}`:

```{r, eval = F}
path_default_nix <- tempdir()

rix(
  r_ver = "4.5.0",
  r_pkgs = c("dplyr"),
  system_pkgs = NULL,
  git_pkgs = list(
    package_name = "BPCells",
    repo_url = "https://github.com/bnprks/BPCells/r",
    commit = "16faeade0a26b392637217b0caf5d7017c5bdf9b"
  ),
  ide = "none",
  project_path = ".",
  overwrite = TRUE,
  print = TRUE
)
```


## Installing local archives

It is also possible to install packages from a local `tar.gz` file. For this,
place the package in the same folder where the `default.nix` will be generated,
and write something like this:

```{r, eval = F}
rix(
  r_ver = "4.3.1",
  local_r_pkgs = c("chronicler_0.2.1.tar.gz", "knitr_1.43.tar.gz"),
  overwrite = TRUE
)
```

This assumes that both `chronicler_0.2.1.tar.gz` and `knitr_1.43.tar.gz` have
been downloaded beforehand. If you want to include a local package that you are
developing, make sure to first build it using `devtools::build()` to build the
`.tar.gz` archive, but if you can, consider uploading the source code to your
package on GitHub and installing it from GitHub instead.

## Installing Python packages

It is also possible to add Python packages to an environment, by passing a list
of two elements to the `py_conf` argument of `rix()`. This list needs to first
specify a Python version, and then an atomic character vector of Python
packages:

```{r, eval = F}
rix(
  date = "2025-02-17",
  r_pkgs = "ggplot2",
  py_conf = list(
    py_version = "3.12", 
    py_pkgs = c("polars", "great-tables")
  ),
  overwrite = TRUE
)
```

This will install Python 3.12, the `polars` and `great-tables` packages, but also
`ipykernel` and `pip` to ensure the Python interpreter works correctly with IDEs 
such as Positron.

## Installing Python packages not available via nixpkgs
Since many Python packages are not available via nixpkgs, you may need to install
them manually. This could be done by using `uv` (which is 10-100x faster than pip
and includes a lock file). The rix call needs to include `uv` while the python installation
could be handled by `uv`.


```{r, eval = F}
rix(
  date = "2025-02-17",
  r_pkgs = "ggplot2",
  system_pkgs = c("uv"),
  overwrite = TRUE
)
```

You then need to manually adjust the resulting `default.nix` with a shellHook and specify your Python packages
in `requirements.txt`.

```nix
...
  shell = pkgs.mkShell {
    LOCALE_ARCHIVE = if pkgs.system == "x86_64-linux" then "${pkgs.glibcLocales}/lib/locale/locale-archive" else "";
    LANG = "en_US.UTF-8";
    LC_ALL = "en_US.UTF-8";
    LC_TIME = "en_US.UTF-8";
    LC_MONETARY = "en_US.UTF-8";
    LC_PAPER = "en_US.UTF-8";
    LC_MEASUREMENT = "en_US.UTF-8";
    
    buildInputs = [  rpkgs   system_packages   ];
    shellHook = ''
      if [ ! -f pyproject.toml ]; then
        uv init --python 3.13.5
      fi
      uv add --requirements requirements.txt
      # Create alias so 'python' uses uv's environment
      alias python="uv run python"
    '';
  }; 
in
  {
    inherit pkgs shell;
  }
```

After running `nix-shell`, `uv` should initalize a Python project with the specified Python version
and install the packages listed in `requirements.txt`. This will take place each time `nix-shell` is called,
however this will be cached and not installed each time.

## Installing Julia packages

Julia support is included as an experimental feature because Nix’s native
handling of Julia is still a work in progress. At the moment, some transitive
dependencies may fail to install. In many cases, explicitly adding those
dependencies by hand will fix the problem; in other cases, it may not. If you
run into issues when installing Julia packages as part of your environment
build, please let us know by opening an issue.

Additionally, date‑specific snapshots of the Julia registry only become
available starting on the 2025-05-20. If you choose an earlier date, Nixpkgs
will fall back to whatever registry snapshot was bundled at that time—which will
be an older version.

To add Julia packages, you can use the `jl_conf` argument of `rix()`. This list
needs to first specify a Julia version, and then an atomic character vector of
packages:

```{r, eval = F}
rix(
  date = "2025-05-19",
  r_pkgs = "ggplot2",
  jl_conf = list(
    jl_version = "1.10",
    jl_pkgs = c("TidierData", "GLM")
  ),
  overwrite = TRUE
)
```

This will install Julia 1.10, the `TidierData` and `GLM` packages.

## Converting from an renv.lock file

`{rix}` also includes an `renv2nix()` function that converts an `renv.lock`
file into a valid Nix expression. Read the vignette `vignette("f-renv2nix")`
to learn more.

## A complete example

This example shows how to install packages from CRAN, from the CRAN archives and
from GitHub:

```{r, eval = F}
path_default_nix <- tempdir()

rix(
  r_ver = "4.2.1",
  r_pkgs = c("dplyr", "janitor", "AER@1.2-8"),
  git_pkgs = list(
    list(
      package_name = "housing",
      repo_url = "https://github.com/rap4all/housing/",
      commit = "1c860959310b80e67c41f7bbdc3e84cef00df18e"
    ),
    list(
      package_name = "fusen",
      repo_url = "https://github.com/ThinkR-open/fusen",
      commit = "d617172447d2947efb20ad6a4463742b8a5d79dc"
    )
  ),
  ide = "none",
  project_path = path_default_nix,
  overwrite = TRUE
)
```

```
#> # This file was generated by the {rix} R package v0.7.1 on 2024-07-01
#> # with following call:
#> # >rix(r_ver = "79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a",
#> #  > r_pkgs = c("dplyr",
#> #  > "janitor",
#> #  > "AER@1.2-8"),
#> #  > git_pkgs = list(list(package_name = "housing",
#> #  > repo_url = "https://github.com/rap4all/housing/",
#> #  > commit = "1c860959310b80e67c41f7bbdc3e84cef00df18e"),
#> #  > list(package_name = "fusen",
#> #  > repo_url = "https://github.com/ThinkR-open/fusen",
#> #  > commit = "d617172447d2947efb20ad6a4463742b8a5d79dc")),
#> #  > ide = "none",
#> #  > project_path = path_default_nix,
#> #  > overwrite = TRUE)
#> # It uses nixpkgs' revision 79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a for reproducibility purposes
#> # which will install R version 4.2.1.
#> # Report any issues to https://github.com/ropensci/rix
#> let
#>  pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/79b3d4bcae8c7007c9fd51c279a8a67acfa73a2a.tar.gz") {};
#>  
#>   rpkgs = builtins.attrValues {
#>     inherit (pkgs.rPackages) 
#>       dplyr
#>       janitor;
#>   };
#>  
#>   git_archive_pkgs = [
#>     (pkgs.rPackages.buildRPackage {
#>       name = "housing";
#>       src = pkgs.fetchgit {
#>         url = "https://github.com/rap4all/housing/";
#>         rev = "1c860959310b80e67c41f7bbdc3e84cef00df18e";
#>         sha256 = "sha256-s4KGtfKQ7hL0sfDhGb4BpBpspfefBN6hf+XlslqyEn4=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages) 
#>           dplyr
#>           ggplot2
#>           janitor
#>           purrr
#>           readxl
#>           rlang
#>           rvest
#>           stringr
#>           tidyr;
#>       };
#>     })
#> 
#> 
#>     (pkgs.rPackages.buildRPackage {
#>       name = "fusen";
#>       src = pkgs.fetchgit {
#>         url = "https://github.com/ThinkR-open/fusen";
#>         rev = "d617172447d2947efb20ad6a4463742b8a5d79dc";
#>         sha256 = "sha256-TOHA1ymLUSgZMYIA1a2yvuv0799svaDOl3zOhNRxcmw=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages) 
#>           attachment
#>           cli
#>           desc
#>           devtools
#>           glue
#>           here
#>           magrittr
#>           parsermd
#>           roxygen2
#>           stringi
#>           tibble
#>           tidyr
#>           usethis
#>           yaml;
#>       };
#>     })
#>  
#>     (pkgs.rPackages.buildRPackage {
#>       name = "AER";
#>       src = pkgs.fetchzip {
#>        url = "https://cran.r-project.org/src/contrib/Archive/AER/AER_1.2-8.tar.gz";
#>        sha256 = "sha256-OqxXcnUX/2C6wfD5fuNayc8OU+mstI3tt4eBVGQZ2S0=";
#>       };
#>       propagatedBuildInputs = builtins.attrValues {
#>         inherit (pkgs.rPackages) 
#>           car
#>           lmtest
#>           sandwich
#>           survival
#>           zoo
#>           Formula;
#>       };
#>     })
#>   ];
#>    
#>   system_packages = builtins.attrValues {
#>     inherit (pkgs) 
#>       R
#>       glibcLocales
#>       nix;
#>   };
#>   
#> in
#> 
#> pkgs.mkShell {
#>   LOCALE_ARCHIVE = if pkgs.system == "x86_64-linux" then  "${pkgs.glibcLocales}/lib/locale/locale-archive" else "";
#>   LANG = "en_US.UTF-8";
#>    LC_ALL = "en_US.UTF-8";
#>    LC_TIME = "en_US.UTF-8";
#>    LC_MONETARY = "en_US.UTF-8";
#>    LC_PAPER = "en_US.UTF-8";
#>    LC_MEASUREMENT = "en_US.UTF-8";
#>
#>   buildInputs = [ git_archive_pkgs rpkgs  system_packages   ];
#>
#> }
```

The next vignette,
`vignette("d2-installing-system-tools-and-texlive-packages-in-a-nix-environment")`,
explains how you can install tools such as text editors, git, Quarto, TexLive
packages, and any other tool available through `nixpkgs` for your development
environments.

## Package installation issues

Some R packages are quite difficult to install: that is usually not an issue for
most users that use either Windows or macOS as their operating systems, because
when calling `install.packages()` a compiled binary gets downloaded from CRAN
and installed in a matter of seconds. On Ubuntu, likely the most popular Linux
distribution, binary packages for R packages are also available via the
[r2u](https://eddelbuettel.github.io/r2u/) repository. However, if you need to
install old packages, these instead will need to be installed from source, as
binaries for old packages are not kept. For most packages, this is not an issue,
but some packages require compilation and this is where issues start.

Nix solves this, because all packages must have their dependencies also
declared, so installing old packages should not be an issue. However, it can
happen that one particular package that you want to install may not build. This
can happen because, even though we spend a lot of time making sure R packages
work flawlessly with Nix, there are many R packages (almost 30'000 between CRAN
and Bioconductor) and there's not many of us (R contributors for Nix). Should
you have trouble installing a package, feel free to open an issue and we'll do
our best to fix it!

We also made sure that old packages would work by backporting many fixes, and
actually building many old versions of popular packages for all the dates
included in `available_dates()`.

Here is the list of packages that were built and tested (but keep in mind that
this list doesn't show all the dependencies of all the packages that also have
to work, and that just because a package isn't listed, doesn't mean it's not
going to work!):

<details>
  <summary>Click to show package list</summary>

```

DBI

R6

RColorBrewer

RCurl

RSQLite

Rcpp

RcppEigen

arrow

askpass

backports

base64enc

bit

bit64

blob

broom

bslib

cachem

callr

cellranger

cli

clipr

collapse

colorspace

cpp11

crayon

curl

data_table

dbplyr

devtools

digest

dplyr

duckdb

evaluate

fansi

farver

fastmap

fontawesome

forcats

fs

gargle

generics

ggplot2

glue

gtable

haven

highr

hms

htmltools

htmlwidgets

httr

icosa

igraph

isoband

jquerylib

jsonlite

kit

knitr

labeling

languageserver

later

lifecycle

lubridate

magrittr

memoise

mime

modelr

munsell

nloptr

openssl

openxlsx

pillar

pkgconfig

prettyunits

processx

progress

promises

ps

purrr

rJava

ragg

rappdirs

readr

readxl

rematch

rematch2

rlang

rmarkdown

rprojroot

rstan

rstudioapi

rvest

sass

scales

selectr

Seurat

sf

shiny

stars

stringi

stringr

sys

systemfonts

terra

textshaping

tibble

tidyr

tidyselect

tidyverse

timechange

tinytex

tzdb

utf8

vctrs

viridisLite

withr

xfun

xlsx

xml2

yaml

zoo

```

</details>
