---
title: "f - Converting renv projects to Nix projects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{e-interactive-use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(rix)
```

## Introduction

`{renv}` is very likely the most popular package to set up per project libraries
of R packages. `{renv}` generates and manages so-called `renv.lock` at the level
of a project, and using these files it is thus possible to have different
versions of the same packages on the same system, without any interference.
However, `{renv}` doesn't snapshot R itself, so different projects with
different libraries of packages will end up using the same version of R, which
could lead to issues, especially if you're trying to restore a old project which
relied on an old version of R. Also, in some cases, it might be impossible to 
restore a project due to incompatibilities at the level of the system-level
dependencies, as these are not managed by `{renv}`.

As explained extensively in this documentation already, Nix handles all the
different pieces of the reproducibility puzzle: versions of R packages, versions
of R, and versions of system-level dependencies are all handled by Nix. In other
words, it's possible to have per project complete development environments that 
are completely reproducible. 

While Nix has been around for 20 years, it was never widely adopted by R users,
who instead have already invested a lot of effort into `{renv}` and Docker.
The goal of `{rix}` is to ease adoption of Nix for R users for new projects,
but there is also a need to convert historical `{renv}` projects into Nix.

This is where `renv2nix()` can help.

## Converting an historical renv project

There are two ways that you can convert an historical project: The recommend way
is to copy the historical `renv.lock` file into a new, empty folder. Do not copy
any of the other generated `{renv}` files nor folders. From there, call
`renv2nix()` like so:

```r
renv2nix(
  renv_lock_path = "path/to/rix_project/renv.lock",
  project_path = "path/to/rix_project"
)
```

This will generate a `default.nix` and `.Rprofile` file for your Nix project.

The other way to achieve this is to point the `renv_lock_path` argument to the
historical `renv.lock` file without copying it to a new folder:

```r
renv2nix(
  renv_lock_path = "path/to/original/renv_project/renv.lock",
  project_path = "path/to/rix_project"
)
```

But this is not the recommended approach. Instead keep the original `renv.lock`
file and the generated `default.nix` file together (ideally with the call to
`renv2nix()` in an R script) and commit everything into version control.

## Starting a new project

If you start a new project, you don't need to use `renv2nix()`, as you could
directly use `rix()` to generate a `default.nix` file. However, you could start
with generating an `renv.lock` file using `renv::snapshot()` and then call
`renv2nix()`, to generate the appropriate `default.nix` file. Because
`renv::snapshot()` does not call `renv::init()` this will not generate any other
file that could interfere with `{rix}`. Doing this could be useful if you have
the habit of starting writing code and then would like to generate a
`default.nix` file later. We don't really recommend working like this and
instead urge you to start from an environment and work from the very beginning
from that environment.

## Caveats

`renv2nix()` has two methods, `"fast"` and `"accurate"`. `"fast"` simply lists
the version of R and R packages, and generates a `default.nix` expression
without trying to match package versions exactly. We believe that this is
acceptable for most use cases. The planned `"accurate"` method will instead try
to do its best to match exact package versions. However, due to how Nix and
`{renv}` handle snapshotting, it will not be possible to match package versions
exactly even with the planned `"accurate"` method. If you absolutely need a very
specific version of a package, there are other ways to achieve this, by
providing for example the required version to the `r_pkgs` argument of `rix()`,
like this: `rix(r_pkgs = "dplyr@0.8.0")`. This however tries to build the pacage
from source, which can fail. If this is the case, don't hesitate to open an
issue. Better handling of individual package versions is planned for a future
release.

As stated previously, don't convert an `renv.lock` file into a `default.nix`
file from a folder that contains the `renv.lock` file and the `{renv}` generated
`.Rprofile` file and `renv/` folder. Instead, work from a new empty folder in
which you copy the `renv.lock` file.
