---
title: "Setting Up and Using Nix and Rix on Linux and Windows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up and using Nix and rix on Linux and Windows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(rix)
```

*This vignette covers Linux- and Windows-specific topics. If you're using
macOS instead, read the `vignette("setting-up-macos")` vignette.*

## Introduction

Nix officially supports only two operating systems: macOS and Linux. Windows
support comes via WSL2, which runs a real Linux distribution under the hood —
so for our purposes, Linux and Windows can be treated as a single case, with
macOS as the separate one. Because Windows relies on WSL2, you'll need WSL2
running before installing Nix.

Note that you don't need Nix installed to use `{rix}` itself: the package can
generate Nix expressions on any system, you just won't be able to *build*
them without Nix. So if you can't install Nix on your machine but already
have R, you can skip ahead to the last section of this vignette to install
`{rix}` on its own.

## Why rix *and* Nix?

To reiterate: `{rix}` will generate valid Nix expressions even on a system
where Nix isn't installed. The catch is that you won't be able to build those
expressions until Nix is actually present.

## Installing Nix

### Windows prerequisites

On Windows, you'll need the Windows Subsystem for Linux 2 (WSL2) to run Nix.
On a recent version of Windows 10 or 11, install it by running the following
as an administrator in PowerShell:

```bash
wsl --install
```

For more details, see [Microsoft's official WSL installation
docs](https://learn.microsoft.com/en-us/windows/wsl/install).

We recommend enabling systemd in Ubuntu WSL2, since this allows users other
than `root` to run Nix. Follow [Microsoft's systemd
guide](https://learn.microsoft.com/en-us/windows/wsl/systemd#how-to-enable-systemd):

```sh
# in WSL2 Ubuntu shell

sudo -i
nano /etc/wsl.conf
```

This opens `/etc/wsl.conf` in nano, a command-line text editor. Add the
following:

```
[boot]
systemd=true
```

Save with CTRL-O, then exit nano with CTRL-X. Back in PowerShell, run:

```bash
wsl --shutdown
```

Then relaunch WSL (Ubuntu) from the Start menu.

Once WSL2 is set up, you can install Nix as usual — we recommend the
Determinate Systems installer, described next.

### Using the Determinate Systems installer

Again, you don't need Nix installed to generate expressions with `{rix}` —
but you do need it if you actually want to build the environments it
defines.

Installing (and uninstalling) Nix is straightforward thanks to the installer
from [Determinate Systems](https://determinate.systems/posts/determinate-nix-installer),
a company that builds services and tools on top of Nix.

Avoid installing Nix through your operating system's package manager.
Instead, open a terminal and run:

```{sh parsermd-chunk-1, eval = FALSE}
curl --proto '=https' --tlsv1.2 -sSf \
     -L https://install.determinate.systems/nix | \
    sh -s -- install --no-confirm --extra-conf "
trusted-users = root $USER
substituters = https://cache.nixos.org https://rstats-on-nix.cachix.org
trusted-public-keys = cache.nixos.org-1:6NCHdD59X431o0gWypbMrAURkbJ16ZPMQFGspcDShjY= rstats-on-nix.cachix.org-1:vdiiVgocg6WeJrODIqdprZRUrhi1JzhBnXv7aWI6+F0="
```

On Windows, if you've decided not to (or can't) enable systemd, append
`--init none` to the command above. See the [Determinate Nix Installer
page](https://github.com/DeterminateSystems/nix-installer) for details.

This installer automatically configures the **rstats-on-nix** Cachix cache,
which provides pre-built binaries for many R packages and significantly
speeds up environment creation. Many thanks to [Cachix](https://www.cachix.org/)
for sponsoring the `rstats-on-nix` cache!

## If you installed Nix another way

If you used the official Nix installer, Lix, or some other method, you'll
need to configure the **rstats-on-nix** Cachix cache yourself.

### Recommended: `setup_cachix()`

The simplest approach is to let `{rix}` configure the cache for you.

First, start R from a temporary Nix shell:

```bash
nix-shell -p R rPackages.rix
```

Then run:

```r
rix::setup_cachix()
```

This updates `~/.config/nix/nix.conf`.

You'll also need to add yourself as a trusted user:

```bash
echo "trusted-users = root $USER" | sudo tee -a /etc/nix/nix.conf && sudo systemctl restart nix-daemon
```

(Note: `/etc/nix/nix.custom.conf` is a Determinate Systems-specific
mechanism, included by their managed `nix.conf` so that your customizations
survive installer updates. Since this section covers non-Determinate
installs, edit `/etc/nix/nix.conf` directly instead.)

If you later see warnings like `ignoring untrusted substituter`, it usually
means this step was skipped.

### Alternative: the `cachix` client

You can instead configure the cache manually with the `cachix`
command-line client:

```bash
nix-env -iA cachix -f https://cachix.org/api/v1/install
cachix use rstats-on-nix
```

### Optional: add Mischko Heming's binary cache

Once the `rstats-on-nix` cache is configured, you can also enable Mischko
Heming's binary cache — see [Using Mischko Heming's binary
cache](#using-mischko-hemings-binary-cache) below.

### NixOS users

On NixOS, Nix configuration is managed declaratively through your system
configuration, so neither `setup_cachix()` nor `cachix use` will work.
Instead, add the cache to your `configuration.nix`:

```nix
nix.settings = {
  substituters = [
    "https://cache.nixos.org"
    "https://rstats-on-nix.cachix.org"
  ];
  trusted-public-keys = [
    "cache.nixos.org-1:6NCHdD59X431o0gWypbMrAURkbJ16ZPMQFGspcDShjY="
    "rstats-on-nix.cachix.org-1:vdiiVgocg6WeJrODIqdprZRUrhi1JzhBnXv7aWI6+F0="
  ];
};
```

If you use Home Manager, add this to your home configuration instead:

```nix
nix.settings = {
  substituters = [
    "https://rstats-on-nix.cachix.org"
  ];
  trusted-public-keys = [
    "rstats-on-nix.cachix.org-1:vdiiVgocg6WeJrODIqdprZRUrhi1JzhBnXv7aWI6+F0="
  ];
};
```

Then rebuild your system with `sudo nixos-rebuild switch` (or `home-manager
switch` for Home Manager).

## Using Mischko Heming's binary cache

To use the [binary cache](https://github.com/mihem/attic) by [Mischko
Heming](https://mheming.com), add this to `~/.config/nix/nix.conf`:

```text
extra-substituters = https://osmzhlab.uni-muenster.de:4949/r-packages
extra-trusted-public-keys = r-packages:Op7Q3XME8az4XNcP1clupGw4ZbuaguBw+sUziweqpTY=
```

On NixOS, add this to your `configuration.nix` instead:

```nix
{
  nix.settings = {
    extra-substituters = [
      "https://osmzhlab.uni-muenster.de:4949/r-packages"
    ];

    extra-trusted-public-keys = [
      "r-packages:Op7Q3XME8az4XNcP1clupGw4ZbuaguBw+sUziweqpTY="
    ];
  };
}
```

On NixOS, rebuild with `sudo nixos-rebuild switch` (or `home-manager switch`
for Home Manager). For regular Nix installations and flake-based projects,
the settings take effect on the next Nix operation.

## A note on disk space

Once Nix is installed on Linux, everything it builds gets stored under the
`/nix` directory on the root partition. Since this partition is often
small, and complete development environments built with Nix can take up
substantial space, we recommend mounting `/nix` on a partition with more
room (for example, a secondary hard drive) if space is tight. To do this,
edit `/etc/fstab` and add the following line at the end:

```
/home/path_to/nix /nix none bind 0 0
```

This maps `/nix` to `/home/path_to/nix`, which can live on a larger
partition. If your root partition already has enough space, you can skip
this step.

## Installing rix

With Nix installed and the binary caches configured, the last step is
getting `{rix}` itself onto your system. How you do that depends on whether
R is already installed:

### Case 1: you don't have R installed and want to install it via Nix

If you have Nix installed but not R, you have two options. You could install
R the usual way for your OS and then install `{rix}` on top of it — from
there, generate project-specific expressions and build them. Or you could
install R through Nix directly. Running the following in a terminal drops
you into an interactive R session you can use right away:

```
nix-shell -p R rPackages.rix
```

Or, if you'd prefer the development version of `{rix}`:

```
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/master/inst/extdata/default.nix)"
```

This should launch an R session directly in your terminal. From there you
can run something like:

```{r, eval = F}
library(rix)

rix(
  r_ver = "4.4.2",
  r_pkgs = c("dplyr", "ggplot2"),
  system_pkgs = NULL,
  git_pkgs = NULL,
  ide = "none",
  project_path = ".",
  overwrite = TRUE
)
```

This generates a `default.nix`, which you can then use to build an
environment with R, `{dplyr}`, and `{ggplot2}`. To add more packages later,
rerun the command with the additional packages listed in `r_pkgs`. Be
careful: if a `default.nix` already exists in your working directory,
running `rix()` with `overwrite = TRUE` will overwrite it — so make sure
you're using version control to avoid unpleasant surprises.

For more on managing project-specific `default.nix` files, see the
`vignette("installing-r-packages")` and `vignette("installing-system-tools")`
vignettes.

You can also include `{rix}` itself in your project-specific environments by
generating a `default.nix` like this:

```{r, eval = F}
rix(
  r_ver = "latest-upstream",
  r_pkgs = NULL,
  git_pkgs = list(
    package_name = "rix",
    repo_url = "https://github.com/ropensci/rix",
    commit = "76d1bdd03d78589d399b4b9d473ecde616920a82"
  ),
  ide = "none",
  project_path = ".",
  overwrite = TRUE
)
```

Update the commit to a more recent one and adjust `project_path` as needed.

### Case 2: you already have R installed via your OS's package manager

*Note: we recommend managing all your R versions through Nix rather than
mixing a system-wide R installation with Nix-managed R shells.*

If R is already installed on your system through the usual channels (i.e.,
not via Nix), you can still install `{rix}` as you normally would:

```{r parsermd-chunk-2-0, eval = FALSE}
install.packages("rix")
```

Or via r-universe:

```{r parsermd-chunk-2, eval = FALSE}
install.packages("rix", repos = c(
  "https://ropensci.r-universe.dev",
  "https://cloud.r-project.org"
))
```

From there, use `{rix}` to generate expressions as described in the next
vignette, `vignette("project-environments")`.

## Configuring an IDE

Next, we recommend continuing with `vignette("installing-r-packages")` and
`vignette("installing-system-tools")`, followed by
`vignette("configuring-ide")`, which walks through setting up your editor to
work with Nix shells effectively.

## Other "Nix"es

Several other implementations of the Nix package manager exist, if you're
feeling adventurous — but for now, we recommend sticking with Nix itself.
If you're curious, check out [Lix](https://lix.systems/install/)!
