---
title: "z - Advanced topic: Handling packages with remote dependencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{z-advanced-topic-handling-packages-with-remote-dependencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(rix)
```

## Introduction

Packages published on CRAN must have their dependencies on either CRAN or
Bioconductor, but not on GitHub. However, there are many packages available on
GitHub that never get published on CRAN, and some of these packages may even
depend on other packages that are also only available on GitHub. `{rix}` makes
it possible to install packages from GitHub and if these packages have
dependencies that are also on GitHub, these also get correctly added to the
generated `default.nix`.

There are however certain caveats you should be aware of.

## The {lookup} package

As an example we are going to use the
[{lookup}](https://github.com/jimhester/lookup) package which has only been
released on GitHub. [Here is the
repository](https://github.com/jimhester/lookup). This package comes with the
`lookup()` function which makes it possible to check the source code of any
function from a loaded package, even if the source of that function is in C or
Fortran. To create a reproducible development environment that makes
[{lookup}](https://github.com/jimhester/lookup) available to you, you could use
the following `rix::rix()` call:

```{r, eval = F}
path_default_nix <- tempdir()

rix(
  r_ver = "latest-upstream",
  r_pkgs = NULL,
  system_pkgs = NULL,
  git_pkgs = list(
    package_name = "lookup",
    repo_url = "https://github.com/jimhester/lookup/",
    commit = "eba63db477dd2f20153b75e2949eb333a36cccfc"
  ),
  ide = "none",
  project_path = path_default_nix,
  overwrite = TRUE,
  print = TRUE
)
```

This will generate the following `default.nix`:

```
let
 pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/ceea3d99cd55e2b343295eea205b19ecc47fe888.tar.gz") {};
  
    httr2 = (pkgs.rPackages.buildRPackage {
      name = "httr2";
      src = pkgs.fetchgit {
        url = "https://github.com/r-lib/httr2";
        rev = "c249b90d5c72d91ad5fab0660904710c56fd4568";
        sha256 = "sha256-x/q2oOqseencAFro73Dxoa2CuyiXnm5NNT1K77HgmLE=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages) 
          cli
          curl
          glue
          lifecycle
          magrittr
          openssl
          R6
          rappdirs
          rlang
          vctrs
          withr;
      };
    });

    gh = (pkgs.rPackages.buildRPackage {
      name = "gh";
      src = pkgs.fetchgit {
        url = "https://github.com/gaborcsardi/gh";
        rev = "HEAD";
        sha256 = "sha256-0ah0s9vvrJQcnOt6Be4oKIo+qD6OAiTTBZ7Upv7Vh9M=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages) 
          cli
          gitcreds
          glue
          ini
          jsonlite
          lifecycle
          rlang;
      } ++ [ httr2 ];
    });


    highlite = (pkgs.rPackages.buildRPackage {
      name = "highlite";
      src = pkgs.fetchgit {
        url = "https://github.com/jimhester/highlite";
        rev = "767b122ef47a60a01e1707e4093cf3635a99c86b";
        sha256 = "sha256-lkWMlAi75MYxiBUYnLwxLK9ApXkWanA4Mt7g4qtLpxM=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages) 
          Rcpp
          BH;
      };
    });


    memoise = (pkgs.rPackages.buildRPackage {
      name = "memoise";
      src = pkgs.fetchgit {
        url = "https://github.com/hadley/memoise";
        rev = "58d39726de141fefd235557a33e6478f76b0ad7f";
        sha256 = "sha256-y3UGutqIFFZn3z5dJzkzs/Fcdwc/+h+5iWofUP5NYic=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages) 
          digest;
      };
    });

    lookup = (pkgs.rPackages.buildRPackage {
      name = "lookup";
      src = pkgs.fetchgit {
        url = "https://github.com/jimhester/lookup/";
        rev = "eba63db477dd2f20153b75e2949eb333a36cccfc";
        sha256 = "sha256-arl7LVxL8xGUW3LhuDCSUjcfswX0rdofL/7v8Klw8FM=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages) 
          Rcpp
          codetools
          crayon
          rex
          jsonlite
          rstudioapi
          withr
          httr;
      } ++ [ highlite gh memoise ];
    });
    
  system_packages = builtins.attrValues {
    inherit (pkgs) 
      glibcLocales
      nix
      R;
  };
  
in

pkgs.mkShell {
  LOCALE_ARCHIVE = if pkgs.system == "x86_64-linux" then "${pkgs.glibcLocales}/lib/locale/locale-archive" else "";
  LANG = "en_US.UTF-8";
   LC_ALL = "en_US.UTF-8";
   LC_TIME = "en_US.UTF-8";
   LC_MONETARY = "en_US.UTF-8";
   LC_PAPER = "en_US.UTF-8";
   LC_MEASUREMENT = "en_US.UTF-8";

  buildInputs = [ lookup   system_packages   ];
  
}
```

as you can see, several other packages hosted on GitHub were added
automatically. This is because these were listed as remote dependencies in
`{lookup}`'s `DESCRIPTION` file:

```
Remotes:
    jimhester/highlite,
    gaborcsardi/gh,
    hadley/memoise
```

`{rix}` uses the GitHub API to fetch the commits of these remote packages and
will attempt to select the commit of the remote packages, whose date is closest
to (always before, never after) the date of the commit provided by the user.

## Caveats

### Fixed commits

`{rix}` is able to fetch a fixed commit specified in DESCRIPTION files using a 
commit hash. This also works in a short form. For example `ropensci/rix@88bb24f`
in `Remotes` in the DESCRIPTION file will work.
However, if a branch name (e.g. `ropensci/rix@available_dates`), tag (e.g. `ropensci/rix@v.0.8.0`), or pull request number
(e.g. `ropensci/rix#100`) is specified, `{rix}` will ignore this and will try to fetch the closest commit (see above).

### Duplicated packages

Let's  say package `A` depends on package `B` and `C`, but package `B` also depends on package `C`. Then 
this would usually result in a duplicate entry of package `C` in the `default.nix` file. 
To avoid this, `{rix}` will cache all packages that have been fetched from GitHub and will not
fetch them again if they are already in the cache. This also avoids unnecessary API calls to GitHub.
If you would like to disable this feature, you can set `ignore_remotes_cache = TRUE` in the `rix()` call.
Be aware that you need to manually remove the duplicates from the `default.nix` file before `nix-build` will work.

Another issue, which is closely related to this problem, is getting the "correct" commit of the dependency.
Let's say the commit of package `A` that you specify in the `rix()` call was committed on the 1st of January 2025.
Then `{rix}` will try to find the commit, whose date is closest to the 1st of January 2025, for package `B` and `C`.
Let's say this is the 1st November 2024 for package `B` and the 1st December 2024 for package `C`. 
If you use `ignore_remotes_cache = TRUE`, then `{rix}` will fetch commits of these dates.
However, if you disable this feature, then `{rix}` will fetch package `C` twice, once for the commit of the 1st December 2024 as before
(closest date that is earlier than the commit date of package `A`) and once for a date
before the 1st November 2024 (closest date that is earlier than the commit date of package `B`).
You should be aware of this because if you are unlucky the default option `ignore_remotes_cache = FALSE` might result in a solution that does not work.


### Authenticating to GitHub

Note: `{rix}` will only be able to fetch a maximum of 1000 commits. That means
if the commit you provided is rather old or if the package has many commits, this
will fail and fallback to the `HEAD` of the repository. We have not observed this
in practice yet, but if this happens you should see a message like this:

```
Failed to get commit date for <<< satijalab/seurat-wrappers >>> No commits found before or on the target date
Falling back to <<< HEAD >>>
```

If you don't have a Github Personal Acccess Token set up, it will show:

```
When fetching the commit date from GitHub from <<< ropensci/rix >>>, no GitHub Personal Access Token found.
Please set GITHUB_PAT in your environment.
Falling back to unauthenticated API request.
```

This should still work until you hit the API limit, in which case you will see
messages, such as this one:

```
Failed to get commit date for <<< satijalab/seurat-wrappers >>> API request failed with status code 403.
Falling back to <<< HEAD >>>
```

The status code 403 tells you that there is a credentials error (e.g., because
you hit the API limit). This is why we highly recommend that you set up a GitHub
Personal Access Token: this will allow `{rix}` to perform authenticated calls to
the API, meaning that `{rix}` will be able to get up to 5000 API calls per hour,
instead of 60, dramatically improving the odds of success. To set a GitHub PAT
for the current session, simply run:

```
Sys.setenv(GITHUB_PAT="YOUR_TOKEN")
```

Of course, you need to create a [GitHub PAT
first](https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens#creating-a-personal-access-token-classic).
If don't want to always have to set the `GITHUB_PAT` environment variable, you
can also use a more [long term solution](https://gitcreds.r-lib.org/index.html).
For more information on GitHub Tokens, please also refer to [Happy Git and GitHub for the useR](https://happygitwithr.com/https-pat). For permanent storage, also refer
to [this stackoverflow question](https://stackoverflow.com/a/5343146/21417317).

Note: If you use `{gitcreds}`, you still need to set the `GITHUB_PAT` variable explicitly for each session before calling `{rix}`.
To avoid exposing your PAT in your code, you could use:

```
my_token <- gitcreds::gitcreds_get()$password
Sys.setenv(GITHUB_PAT = my_token)
```

However, you might still be getting the following message, even after
correctly configuring a GitHub PAT:

```
Failed to get closest commit for gaborcsardi/gh: Failed to download commit data
Argument 'url' must be string..
Falling back to <<< HEAD >>>
```

In this particular case, it might be due to the fact that the url
`https://github.com/gaborcsardi/gh` now points to `https://github.com/r-lib/gh`.
In instances like this, we recommend to manually change the url and the revision
from `"HEAD"` to an actual commit. If you don't do this, this means that if
these repositories are being actively worked on, rebuilding these environments
will actually pull another version of these packages. So we very highly
recommend to edit the `default.nix`, and replace mentions of `HEAD` (if any)
with an actual commit. For example, edit this:

```
gh = (pkgs.rPackages.buildRPackage {
  name = "gh";
  src = pkgs.fetchgit {
    url = "https://github.com/gaborcsardi/gh";
    rev = "HEAD";
    sha256 = "sha256-0ah0s9vvrJQcnOt6Be4oKIo+qD6OAiTTBZ7Upv7Vh9M=";
  };
  propagatedBuildInputs = builtins.attrValues {
    inherit (pkgs.rPackages)
      cli
      gitcreds
      glue
      ini
      jsonlite
      lifecycle
      rlang;
  } ++ [ httr2 ];
});
```

to this:

```
gh = (pkgs.rPackages.buildRPackage {
  name = "gh";
  src = pkgs.fetchgit {
    url = "https://github.com/gaborcsardi/gh";
    rev = "27db16cf363dc";
    sha256 = ""; # <- You will need to try to build the expression once, and then
  };             # <- put the sha256 that nix-build returns
  propagatedBuildInputs = builtins.attrValues {
    inherit (pkgs.rPackages)
      cli
      gitcreds
      glue
      ini
      jsonlite
      lifecycle
      rlang;
  } ++ [ httr2 ];
});
```

Finally, and if instead the remotes are listed like this:

```
Remotes:
    jimhester/highlite@abc123,
    gaborcsardi/gh@def123,
    hadley/memoise@ghi123
```

then the listed commits will be used, which will make sure that the build
process is reproducible. Only commits can be used, anything else listed there
(such as pull request numbers or release tags) will not work with `{rix}`.

### Consider using dependencies that made it to CRAN

`{highlite}` is a dependency of [{lookup}](https://github.com/jimhester/lookup)
that is only available on GitHub. `{gh}` and `{memoise}` are also listed as
remote dependencies, however, they are also available on CRAN now. This was not
the case at the time when `{lookup}` was written (which was more than 6 years
ago as of 2025). Because they are listed as remote dependencies, they will also
be built from GitHub instead of CRAN.

Here, it is up to you to decide if you want to keep the GitHub version of these
packages, or if you should instead include the released CRAN version. Depending
on what you want to do, going for the CRAN release of the packages might be
advisable. For example in this case, trying to build this expression will not
work.

This is because `{httr2}` is a package that needs to be compiled from source and
which needs some Nix-specific fixes applied to its source code for it to build
successfully. Installing the version provided by `nixpkgs`, which builds upon
the released CRAN version will succeed however. To do so, change the
`default.nix` manually to this (essentially remove the definition of `{httr2}`
and put it as a `propagatedBuildInput` to `{gh}`):

```
let
 pkgs = import (fetchTarball "https://github.com/NixOS/nixpkgs/archive/6a25f33c843a45b8d67ba782b6782973a7265774.tar.gz") {};

    gh = (pkgs.rPackages.buildRPackage {
      name = "gh";
      src = pkgs.fetchgit {
        url = "https://github.com/gaborcsardi/gh";
        rev = "HEAD";
        sha256 = "sha256-VpxFIfUEk0PudytQ3boMhEJhT0AnelWkSU++WD/HAyo=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages)
          cli
          gitcreds
          glue
          httr2 # <- httr2 is now declared here, so it's the CRAN version
          ini
          jsonlite
          lifecycle
          rlang;
      };
    });

    highlite = (pkgs.rPackages.buildRPackage {
      name = "highlite";
      src = pkgs.fetchgit {
        url = "https://github.com/jimhester/highlite";
        rev = "767b122ef47a60a01e1707e4093cf3635a99c86b";
        sha256 = "sha256-lkWMlAi75MYxiBUYnLwxLK9ApXkWanA4Mt7g4qtLpxM=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages)
          Rcpp
          BH;
      };
    });

    memoise = (pkgs.rPackages.buildRPackage {
      name = "memoise";
      src = pkgs.fetchgit {
        url = "https://github.com/hadley/memoise";
        rev = "58d39726de141fefd235557a33e6478f76b0ad7f";
        sha256 = "sha256-y3UGutqIFFZn3z5dJzkzs/Fcdwc/+h+5iWofUP5NYic=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages)
          digest;
      };
    });

    lookup = (pkgs.rPackages.buildRPackage {
      name = "lookup";
      src = pkgs.fetchgit {
        url = "https://github.com/jimhester/lookup/";
        rev = "eba63db477dd2f20153b75e2949eb333a36cccfc";
        sha256 = "sha256-arl7LVxL8xGUW3LhuDCSUjcfswX0rdofL/7v8Klw8FM=";
      };
      propagatedBuildInputs = builtins.attrValues {
        inherit (pkgs.rPackages)
          Rcpp
          codetools
          crayon
          rex
          jsonlite
          rstudioapi
          withr
          httr;
      } ++ [ highlite gh memoise ];
    });

  system_packages = builtins.attrValues {
    inherit (pkgs)
      glibcLocales
      nix
      R;
  };

in

pkgs.mkShell {
  LOCALE_ARCHIVE = if pkgs.system == "x86_64-linux" then "${pkgs.glibcLocales}/lib/locale/locale-archive" else "";
  LANG = "en_US.UTF-8";
   LC_ALL = "en_US.UTF-8";
   LC_TIME = "en_US.UTF-8";
   LC_MONETARY = "en_US.UTF-8";
   LC_PAPER = "en_US.UTF-8";
   LC_MEASUREMENT = "en_US.UTF-8";

  buildInputs = [ lookup   system_packages   ];

}
```

In this manually edited expression, `{httr2}` will now build successfully
because Nix is instructed to build the CRAN version by applying [this
fix](https://github.com/NixOS/nixpkgs/blob/7b87fced8bc525d466c7258a042bd12ea336a3c6/pkgs/development/r-modules/default.nix#L1817)
which was added there by packagers and maintainers of the R programming language
for `nixpkgs` (it is exactly the same if you tried to install `{httr2}` from
GitHub on Windows: you would need to build it from source and thus make sure
that you have the required system-level dependencies to build it. Instead, it is
easier to install a pre-compiled binary from CRAN).

## Conclusion

In conclusion, `{rix}` makes it easier to build packages from GitHub which have
themselves dependencies hosted on GitHub, you should however make sure that the
expression that is generated uses fixed commits instead of `HEAD` for the
packages being built from GitHub (if `{rix}` wasn't able to automatically fetch
the right commit for you), and you should also decide if you want to use the
version of a packages hosted on GitHub instead of the CRAN release. These are
decisions that `{rix}` cannot take for you.


