---
title: "z - Advanced topic: Rolling out your own binary cache"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{z-advanced-topic-rolling-out-your-own-binary-cache}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(rix)
```

*We recommend you first read `vignette("z-bleeding_edge")` before this one.*

## Introduction

At its core, the Nix package manager installs software from source, which can
take a very long time. For example, installing R or RStudio from source could
take quite some time, depending on your hardware. Some R packages also must be
compiled, so depending on the environment you need, this could take several
hours to build. In practice however, this is rarely the case as most, if not all
packages get pre-built and made available through NixOS's [public
cache](https://cache.nixos.org/). The Nix package manager first checks if what
you need is in the public cache, and if yes, it downloads a binary from there
instead of building it on your computer. This means that building an environment
is just a matter of waiting for packages to download.

However, in some cases, you might need to use packages that have not been
pre-built and cached yet, for example if you use `rix(r_ver = "bleeding_edge")`
or rix(r_ver = "frozen_edge")` to generate environments with bleeding edge
packages. Building these environments could potentially take quite long, because
these packages need to be built from source on your computer.

This vignette will explain how you can cut down on building times by building
the environment once, and then cache it, so you don’t need to rebuild it on new
machines. But before rolling out your own cache, try using ours as explained in
`vignette("z-bleeding_edge")`, and if our cache didn’t cover your needs, set up
your own. This vignette explains how.

## Using Github Actions to build the environment

Before creating a cache to hold the binaries for your development environment,
you first need to build the environment once. You can build the environment on
your computer, or on Github Actions. The advantage of using Github Actions is
that you can automate the process of building and pushing the binaries each time
you change the definition of the environment.

For example, imagine you have the following `generate_env.R` file for a project,
which you version on Github:

```
library(rix)

rix(r_ver = "bleeding_edge",
    r_pkgs = c("dplyr", "ggplot2"),
    system_pkgs = NULL,
    git_pkgs = NULL,
    tex_pkgs = NULL,
    ide = "rstudio",
    project_path = ".")
```

if you add packages to it, or re-run it, you’ll end up with a new
`default.nix` file, and so you will need to rebuild the environment. Again,
depending on the packages you include, this could take quite some time to build.
To generate a Github Actions workflow file that will build the environment on
Github Actions, run `rix::ga_cachix(cache_name = "", path = "")` where
`cache_name` is the name of the cache you made on Cachix, and `path` is the path
to the `default.nix` file generated by `generate_env.R`. By default, the
environment gets rebuilt every time you push changes to the `master/main` branch
of your repository, but you can also re-build the environment periodically, by
changing these lines:

```
on:
  push:
    branches: [ master, main ]
```

to these:

```
on:
  push:
    branches: [ master, main ]
  schedule:
    - cron: '30 0 * * *'
```

By using [cron syntax](https://en.wikipedia.org/wiki/Cron#Overview) you can
specify how often you want the environment to be re-built. This can be useful
if you need to develop against the current state of CRAN every day (for
instance, for package development).

## Using your cache

Whether you decide to build the packages on Github or locally, to then use your
cache, you need to open an account on [Cachix](https://www.cachix.org/). The
free tier includes 5GB of space, which is more than enough for several
development environments. Once your account is done, create a personal auth
token so that the Github Actions workflow (or your computer, if building
locally) can authenticate to your Cachix account. Then, copy this token, and go
to your Github repository’s settings, then *Secrets and variables > Actions* and
add a new repository secret. Copy the token in the `Secret` field and name the
secret `CACHIX_AUTH`. Now the action can authenticate to Cachix and push the
binaries it builds! If building locally, simply run `cachix authtoken <TOKEN>`
in your terminal before building (follow the instructions on Cachix website to
learn how to push the binaries afterwards).

To use your personalized cache on your computers, run the following commands on
your computer. First, install the Cachix client:

```
nix-env -iA cachix -f https://cachix.org/api/v1/install
```

then use the cache:

```
cachix use your-cache-name
```

Anyone can pull binaries from your cache, so if you work in a team, you can
ensure everyone can benefit from it. You can also use several caches at once,
NixOS's public cache, our `rstats-on-nix` cache, and your own, so your cache
will only end up holding the binaries not found in the other two caches!

Take a look at
[this package’s](https://github.com/ropensci/rix/blob/main/.github/workflows/cachix-dev-env.yml)
repository for an example of how this is done in practice.
